---
title: "CMSA Camp Lectures"
author: "Tess Hinkle"
format: html
editor: visual
---

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

## Lecture One Code : Exploring Data in tidyverse

Using package Lahman for baseball statistics.

```{r}
#| message: false
#| warning: false
require(tidyverse)
require(Lahman)
require(gt)
require(dplyr)
require(mosaic)
```

Getting the basic information of the Batting data set. The function `dim()` provides the number of rows and columns of the data set instead of getting it separately through `nrow()` and `ncol()`.

```{r}
Batting = as_tibble(Batting)
view(Batting)

dim(Batting)
nrow(Batting)
ncol(Batting)
```

Getting an overview of the code. The function `head()` gives the first 6 rows of a data set. The function `str()` gives a look at each variable within the data set and what data type it is such as a character, integer, or factor. The function `summary()` shows the simple statistics of each variable in the data set. To get one particular variable's summary the following can be used: `summary(data_set$variable)`. The function `table()` gives the number of observations per category of a variable.

```{r}
head(Batting)

str(Batting)

#summary(Batting)
summary(Batting$G)

table(Batting$lgID)
```

### Data Wrangling with dplyr

Filtering extracts data from a data set with specific conditions such as the league IDs being those of the MLB leagues. Or more specifically it can be used to get a specific year of the data for a specific baseball team. The function `select()` allows for specific variables to be chosen from a data set.

```{r}

mlb_batting = filter(Batting, lgID %in% c("AL", "NL")) 
# or filter(Batting, lgID == "AL" | lgID == "NL")
nrow(Batting) - nrow(mlb_batting) # difference in row counts

#summary(mlb_batting)

pirates_batting = filter(Batting, yearID == 2022 & teamID == "PIT")
head(pirates_batting, n = 3) # difference in row counts

select(Batting, playerID, G, AB, R, H, HR, BB)

select(Batting, contains("ID"), G, AB, R, H, HR, BB)
```

Using the `mutate()` function allows for a new column/variable to be made using existing variables. The `arrange()` function allows for the data set to be arranged in descending or ascending order based on a particular variable.

```{r}
new_batting = mutate(Batting, batting_avg = H / AB, so_bb_ratio = SO / BB)
head(new_batting, n = 5)

hr_batting = arrange(Batting, desc(HR)) # desc() for descending order
head(hr_batting, n = 3)

arrange(Batting, desc(AB), HR)
```

The pipe operator allows for multiple functions such as `filter()` and `mutate()` to be applied to a data set all at once. The `summarize()` function allows for data to be collapsed into a single summary value such as mean, median, max, min, and correlation. Another similar way to get summary variable is using the `favstats()` function or the `summary()` function. However, the `summarize()` function allows for these summaries to be used a variables. The `group_by()` function collapses groups in the called column.

```{r}
Batting |> 
  filter(yearID == 2022, teamID == "PIT", AB >= 50) |> 
  mutate(batting_avg = H / AB) |> 
  arrange(desc(batting_avg)) |> 
  select(playerID, AB, batting_avg)

#the summarize() function
Batting |> 
  summarize(median_at_bats = median(AB))
Batting |> 
  summarize(cor_ab_hr = cor(AB, HR))

favstats(~AB,data = Batting)

#using the group_by() function
Batting |> 
  filter(yearID %in% 2015:2019) |> 
  group_by(teamID) |> 
  summarize(total_hr = sum(HR),
            total_so = sum(SO),
            total_bb = sum(BB)) |> 
  arrange(desc(total_hr))


```

The count function give the number of observation in each category of a variable. Such as how many times the AL ID is called is 54277

```{r}
Batting |> 
  count(lgID, name = "freq")
```

::: callout-note
Note that there are other ways to get the same thing that `count()` does. One way is the base R way using `table()`. Another longer way would be `Batting |> group_by(lgID) |> summarize(freq = n()) |> ungroup()`
:::

A way to get specific rows is using the `slice()` function. The function `tail()` is the direct opposite of `head()` providing the last six rows of a data set.

```{r}
Batting |> 
  slice(c(1, 99, 101, 500)) #sliced by row index

tail(Batting)

# single-season home run record (top 5)
Batting |> 
  slice_max(HR, n = 5)
```

### Putting it all together

Example 1: Getting the batting stats for each year Example 2: Finding the top 3 years with the most homeruns Example 3: Finding out which years have the best and worst strikeout to walk ratios

```{r}
yearly_batting = Batting |>
  filter(lgID %in% c("AL", "NL")) |>
  group_by(yearID) |>
  summarize(total_hits = sum(H, na.rm = TRUE), #total hits
            total_hr = sum(HR, na.rm = TRUE), #total home runs
            total_so = sum(SO, na.rm = TRUE), #total strikeouts
            total_bb = sum(BB, na.rm = TRUE), #total walks
            total_ab = sum(AB, na.rm = TRUE)) |> #total at bats
  mutate(batting_avg = total_hits / total_ab)

yearly_batting|>
  arrange(desc(total_hr))|>
  slice(1:3)
#or can rewrite as
#yearly_batting |> 
#  slice_max(total_hr, n = 3)

yearly_batting |>
  mutate(so_bb_ratio = total_so/total_bb) |>
  select(-batting_avg) |>
  arrange(so_bb_ratio) |>
  slice(c(1,n()))
```

Making pretty tables. The `gt()` function creates tables that can be exported and look better than the regular tibble tables.

```{r}

yearly_batting |>
  select(yearID, batting_avg) |>
  rename(Year = yearID, `Batting Average` = batting_avg) |>
  arrange(desc(`Batting Average`)) |>
  slice(c(1:3, (n()-2):n())) |>
  gt() |>
  fmt_number(columns = `Batting Average`, decimals = 3)

yearly_batting |>
  select(yearID, batting_avg) |>
  rename(Year = yearID,
         `Batting Average` = batting_avg) |>
  arrange(desc(`Batting Average`)) |>
  slice(c(1:3, (n()-2):n())) |>
  gt(rowname_col = "Year") |>
  # add table header
  tab_header(title = "Best / Worst MLB Seasons by Batting Average") |>
  # create row groups by location 
  tab_row_group(
    label = "Bottom 3 Years",
    rows = 4:6
  ) |>
  tab_row_group(
    label = "Top 3 Years",
    rows = 1:3
  ) |>
  # recolor row group labels
  tab_style(
    style = cell_fill(color = "lightblue"),
    locations = cells_row_groups(groups = "Bottom 3 Years")
  ) |>
  tab_style(
    style = cell_fill(color = "lightpink"),
    locations = cells_row_groups(groups = "Top 3 Years")) |>
  # round "Batting Average" to 3 decimals 
  fmt_number(columns = `Batting Average`, decimals = 3)

```

## Lecture Two Code: Grammar of Graphics and ggplot2

```{r}
#| include: false
require(ggplot2)
require(Lahman)
library(tidyverse)
require(babynames)
theme_set(theme_bw())

yearly_batting <- Batting |>
  filter(lgID %in% c("AL", "NL")) |>
  group_by(yearID) |>
  summarize(total_h = sum(H, na.rm = TRUE),
            total_hr = sum(HR, na.rm = TRUE),
            total_so = sum(SO, na.rm = TRUE),
            total_bb = sum(BB, na.rm = TRUE),
            total_ab = sum(AB, na.rm = TRUE)) |>
  mutate(batting_avg = total_h / total_ab)
```

Continuing with the Baseball batting data set from the Lahman package.

Creating simple scatterplots with `geom_plot`. Can add on function `theme_bw()` to specify what theme is wanted for the graph beyond the base theme, or can use `theme_set()` to set theme for all plots at the beginning of code.

```{r}
yearly_batting |> 
  ggplot() +
  geom_point(aes(x = yearID, y = total_hr))
```

Creating line plots using `geom_point()` and `geom_line()`.

```{r}
#| warning: false
#| message: false
ggplot(data = yearly_batting, aes(x=yearID, y=total_hr)) + geom_point() +
  coord_cartesian() + geom_line() + theme_bw() +
  scale_x_continuous(limits = c(2000,2015)) + #looks at downward trend from 2000-2015
  labs(title = "Yearly Batting for the MLB", y = "Total Home Runs", x = "Year")

ggplot(yearly_batting, aes(x= yearID, y = total_hr)) + geom_point()+
  geom_line() + scale_x_reverse() +scale_y_log10() + theme_bw()

ggplot(yearly_batting, aes(x= yearID, y = total_hr)) + geom_point()+
  geom_line() + geom_smooth() + theme_bw()
```

Creating a complex scatter plot of the rise of homeruns, strikeouts, and walks. The number of Homeruns per year are noted on the y-axis while the number of strikeouts are noted in the scaled color gradient and the number of walks is noted in the size of the points on the scatter plot. Labels are done for every possible variable including a title and caption.

```{r}
ggplot(yearly_batting, aes(x = yearID, y = total_hr)) + 
  geom_point(aes(color = total_so, size = total_bb)) +
  geom_line(color = "darkred", linetype = "dashed") +
  scale_color_gradient(low = "darkblue", high = "gold") +
  labs(
    x = "Year",
    y = "Homeruns",
    color = "Strikeouts",
    size = "Walks",
    title = "The rise of three true outcomes in baseball",
    caption = "Data courtesy of Lahman"
  )
```

The art of pivoting. The function `pivot_longer()` takes data that is in a wide format and transforms it to a long format. It changes the data set to have more columns and less rows. The function `pivot_wider()` is the exact opposite and widens the data set to have more columns and less rows. Typically pivoting longer is more widely used.

```{r}
#| output: false
yearly_batting |> 
  select(yearID, HRs = total_hr, Strikeouts = total_so, Walks = total_bb) |> # renaming while also selecting
  pivot_longer(HRs:Walks, # can also do !yearID (to select everything but yearID)
               names_to = "stat",
               values_to = "val")
```

Faceting allows for multiple graphs to be shown in one pane.The function `facet_wrap()` allows for multiple graphs to be on one pane. Calling `~ variable` within the function allows for the graph to group by that variable for the facet wrap. A `facet_grid()` function is similar to the wrap function but the syntax is slightly different calling the grouped variable before the tilda.

```{r}
yearly_batting |>
  select(yearID, HRs = total_hr, 
         Strikeouts = total_so, Walks = total_bb) |>
  pivot_longer(HRs:Walks, 
               names_to = "stat", 
               values_to = "val") |>
  ggplot(aes(yearID, val)) +
  geom_line(color = "darkblue") +
  geom_point(alpha = 0.8, color = "darkblue") +
  facet_wrap(~ stat, scales = "free_y", ncol = 1) +
  labs(
    x = "Year", 
    y = "Total of statistic",
    title = "The rise of three true outcomes in baseball",
    caption = "Data courtesy of Lahman"
  ) +
  theme_bw(base_size = 20) +
  theme(strip.background = element_blank(),
        plot.title = element_text(hjust = 0.5, 
                                  face = "bold"))
```

## Lecture 3 Code: Data Visualization w/ categorical variable
Working with Tennis shot data from Wimbeldon's women's matches. Data taken from the SCORE sports data repository.

```{r}
#| include: false
library(tidyverse)
require(janitor)
library(gt)

tennis_shots = read_csv("tennis-w-shots-wim.csv.gz") |>
  # tidy names
  janitor::clean_names() |>
  # remove serves
  filter(shot_type != "serve") |>
  # NA = shot did not produce an outcome (i.e., returned)
  mutate(outcome_type = if_else(is.na(outcome_type),"returned", outcome_type)) |>
  # replace "_" with " " in all character variables 
  mutate(across(where(is.character), ~ str_replace_all(., "_", " "))) |>
  # convert date into ymd format
  mutate(date = ymd(date)) |>
  # random sample of sample 10,000 observations
  dplyr::slice_sample(n = 10000)

tennis_shots = read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/tennis_w_shots_wim.csv") |>
  # filter out NAs for shot_type and outcome_type
  filter(!is.na(shot_type), !is.na(outcome_type)) |>
  # convert characters into factors
  mutate(across(where(is.character), as.factor)) #mutate anytime variable is a character into a factor
```

Getting simple statistics of the data set.
```{r}
head(tennis_shots)

names(tennis_shots) #gives column names

class(tennis_shots$shot_type) #tells that its a factor
levels(tennis_shots$shot_type) #gives the different factor levels

table(tennis_shots$shot_type)
prop.table(table(tennis_shots$shot_type))
```

Plotting categorical variables. Using angle allows for long category names to be viewed easily without the names overlapping. Specifying that the categorical variable is on the y-axis flips the bar chart. A similar way to do this is using `coord_flip()`. Using `geom_label()` allows for the number of observations of each category to be shown. Using `geom_errorbar()` gives errorbars for each bar.

```{r}
ggplot(tennis_shots,aes(x = shot_type)) +
  geom_bar(fill = "lightblue", col = "blue") +
  # angling axis text for readability
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))

tennis_shots |> 
  ggplot(aes(y = shot_type)) +
  geom_bar(fill = "blue") 

tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n)) |> 
  ggplot(aes(x = prop, y = shot_type)) +
  geom_col()  + geom_label(aes(label = n), hjust = 1)

tennis_shots |> 
  count(shot_type) |> 
  mutate(prop = n / sum(n),
         se = sqrt(prop * (1 - prop) / sum(n)),
         lower = prop - 2 * se,
         upper = prop + 2 * se) |> 
  ggplot(aes(x = prop, y = shot_type)) +
  geom_col(fill = "lavender") +
  geom_errorbar(aes(xmin = lower, xmax = upper), 
                color = "blue", 
                width = 0.2, 
                linewidth = 1)
```

Looking into multiple variables at once. Using `fct_relevel()` is able to move specified categories to the front. R goes alphabetically, so if we want to put a specific category towards the front we can relevel.

```{r}
table(tennis_shots$shot_type)

tennis_shots = mutate(tennis_shots, outcome_type = fct_relevel(outcome_type, "unforced error"))
table(tennis_shots$outcome_type)

#two-level table
table("Shot type" = tennis_shots$shot_type, 
      "Outcome type" = tennis_shots$outcome_type)
#another way to create a two-level table
xtabs(~ shot_type + outcome_type, data = tennis_shots)
```

Creating graphs with 2+ variables.

```{r}
tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n, fill = outcome_type)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))

tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n,
             fill = outcome_type)) +
  geom_col(position = "fill") + #changes from count to proportion
  labs(y = "prop") +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))

tennis_shots |>
  count(shot_type, outcome_type) |>
  ggplot(aes(x = shot_type, y = n,
             fill = outcome_type)) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```

Creating proportion tables.

```{r}
# The base `R` way to get a two-way proportion table
tennis_shots |> 
  select(shot_type, outcome_type) |> 
  table() |> 
  prop.table()
```
Creating a heat map of shot types and outcome types
```{r}
tennis_shots |>
  group_by(shot_type, outcome_type) |>
  summarize(
    freq = n(), 
    joint = n() / nrow(tennis_shots)
  ) |> 
  ggplot(aes(x = shot_type, y = outcome_type)) +
  geom_tile(aes(fill = freq), color = "white") + #creates heat map
  geom_text(aes(label = scales::percent(joint, accuracy = 0.01))) +
  scale_fill_gradient2() +
  coord_equal() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1, hjust = 1))
```

## Lecture 4 Code: Data viz w/ quantitative data
Exploring ultra trail racing, specifically looking at the Western States Endurance Run. 

```{r}
#| include: false
require(tidyverse)
library(ggbeeswarm)
library(GGally)
require(ggridges)

runs = read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ultra_running.csv") 
race_stats_2019 = read_csv("https://raw.githubusercontent.com/36-SURE/2026/main/data/ultra_races_2019.csv")

wser = runs |>
  filter(event == "Western States Endurance Run")
```

Getting simple statistics of the western states endurance run data set. The function `sd()` gets the specific standard deviation of a variable from the data set.
```{r}
summary(wser$time_in_hours)

sd(wser$time_in_hours, na.rm = TRUE) #na.rm removes na's
```

Creating a boxplot and histogram looking at the time in hours of race finishes. The histogram shows a bimodal model where a large portion of races finish within the 24 hour mark and another peak right before 30 hours.
```{r}
wser |> 
  ggplot(aes(x = time_in_hours)) +
  geom_boxplot() +
  theme(axis.text.y = element_blank())+
  labs(x= 'Time (hours)')

wser |> 
  ggplot(aes(x = time_in_hours)) +
  geom_histogram(fill = "lightblue", col = "blue")+
  labs(x= "Time (hours)")
```

Creating a densitz plot which is similar to a histogram but is a smooth line rather than bins.
```{r}
wser |> 
  ggplot(aes(x = time_in_hours)) +
  geom_density()+
  labs(x= "Time (hours)")
```

A beeswarm plot to look at each individual data point, shows the bimodal-ness of the data as well as density with how many points are in a specific line. A violin plot is essentially a density plot of a beeswarm plot. A boxplot can be added to a violin plot using the `geom_boxplot()` function as well as the `geom_violin()` function at the same time.
```{r}
wser |> 
  ggplot(aes(x = time_in_hours, y = "")) +
  geom_beeswarm(cex = 1, size = 0L)+
  labs(x= "Time (hours)", y="")

wser |> 
  ggplot(aes(x = time_in_hours, y = "")) +
  geom_violin()+
  labs(x= "Time (hours)", y ="")

wser |> 
  ggplot(aes(x = time_in_hours, y = "")) +
  geom_violin() +
  geom_boxplot(width = 0.25)+
  labs(x= "Time (hours)", y = "")
```

Creating an ECDF plot where we can see the proportion of data under each curve. We can see that 50% of the data is at about 26 hours. We can also see a peak in the data at 24 hours showing how a large amount of the data takes place at that time and then levels out until it increases again. To create an ECDF plot use the function `stat_ecdf()` function.
```{r}
wser |> 
  ggplot(aes(x = time_in_hours)) +
  stat_ecdf()+
  labs(x= "Time (hours)")
```

### Plotting 2D quantitative data
Creating a scatterplot to look at the relationship between elevation gain and average finishing time in hours. Can create a LSRL using the `geom_smooth(method = "lm")` function.
```{r}
race_stats_2019 |> 
  ggplot(aes(x = elevation_gain, y = mean_finish_time)) +
  geom_point(color = "navy", size = 4, alpha = 0.5)+
  scale_x_continuous(labels = scales::label_comma())+
  labs(x = "Elevation Gain (m)", y = "Avg finish time (hrs)")

cor(race_stats_2019$elevation_gain, 
    race_stats_2019$mean_finish_time, 
    use = "complete.obs") # to ensure no missing data

race_stats_2019 |> 
  ggplot(aes(x = elevation_gain, y = mean_finish_time)) +
  geom_point(color = "navy", size = 4, alpha = 0.5)+
  geom_smooth(method = "lm", linewidth = 2)+
  labs(x = "Elevation Gain (m)", y = "Avg finish time (hrs)")

```

To compare multiple variables at a time you can use the GGally package which shows the correlation, density plots, and scatterplot between the variables.
```{r}
race_stats_2019 |> 
  ungroup() |>
  select(elevation_gain, elevation_loss, mean_finish_time, n_finishers) |> 
  ggpairs(aes(alpha=0.3))
```

We can also look at categorical data with quantitative data. The function `case_when()` allows for you to group a quantitative variable into a categorical variable. The boxplot shows how different age groups have different median finishing times. The ECDF plot also shows the percentage of the groups that is within the finishing time. We are able to see how the youngest age group intercept the 50% line prior to 25 hours while the other three groups cross the 50% line after 25 hours.
```{r}
wser <- wser |>
  mutate(age_grp = case_when(age <= 39 ~ "22-39", 
                             age > 39 & age <= 49 ~ "40-49", 
                             age > 49 & age <= 59 ~ "50-59", 
                             age > 59 ~ "60+"))

wser |> 
  ggplot(aes(x = time_in_hours, y = age_grp))+
  geom_boxplot()+
  labs(x = "Time (hours)", y = "Age group")

wser |> 
  ggplot(aes(x = time_in_hours, color = age_grp)) +
  stat_ecdf(linewidth = 1) +
  theme(legend.position = "bottom")+
  scale_color_viridis_d()+
  labs(x = "Time (hours)", color = "Age group")
```

We can also create density plots with separate density plots for each category.
```{r}
wser |> 
  ggplot(aes(x = time_in_hours, color = age_grp)) +
  geom_density(linewidth = 0.8) + 
  theme(legend.position = "bottom") +
  labs(color = "Age group", x = "Time (hours)")

wser |> 
  ggplot(aes(x = time_in_hours, y = age_grp, fill = age_grp)) +
  geom_density_ridges(scale = 1.5)+
  scale_fill_viridis_d()+
  labs(x = "Time (hours)", y = "Age group")+
  theme(legend.position = "none")
```

The same thing can be done with histograms.
```{r}
wser |> 
  ggplot(aes(x = time_in_hours, fill = age_grp)) +
  geom_histogram(bins = 15, col = "black") +
  scale_fill_viridis_d(alpha = 0.6, option= "mako")+
  labs(fill = "Age group", x = "Time (hours)")

wser |>
  ggplot(aes(x = time_in_hours)) +
  geom_histogram(bins = 15, col = "blue", fill = "lightblue") +
  facet_wrap(~ age_grp, nrow = 1)+
  labs(x= "Time (hours)")
```

