---
title: "CMSA Lecture 22"
format: html
editor: visual
---

```{r}
#install.packages(c("DBI", "dittodb", "RSQLite", "nycflights13")) need to install all these
library(DBI)
library(dittodb)
library(RSQLite)

# establishing connection via sqlite
# ":memory:" is a special path that creates an in-memory database
# other database drivers require more details (like user, password, host, port, etc.)
con_air = DBI::dbConnect(RSQLite::SQLite(), ":memory:") 
# creates a database including a few tables from the nycflights13 dataset with con_air connection
dittodb::nycflights13_create_sql(con_air)

DBI::dbListTables(con_air) #gives what tables you have access to
```
Creating tables

```{r}
flights_tbl = tbl(con_air, "flights")
dim(flights_tbl)  #head(flights_tbl, n = 3) 

dist_sum = flights_tbl |>
  summarize(min_year = min(distance), max_year = max(distance))

show_query(dist_sum)
```
SQL Tables as tibble

```{r}
flights_tibble = flights_tbl |> dplyr::collect()

# dim(flights_tibble)
head(flights_tibble, n = 3)

flights_tbl |>
  object.size() |>
  print(units = "Kb")

flights_tibble |>
  object.size() |>
  print(units = "Kb")
```

Different ways to query on quarto

```{r}
flights_tbl |>
  summarize(min_year = min(distance), max_year = max(distance)) # what we previously did

DBI::dbGetQuery(con_air,
                "SELECT MIN(distance) AS min_year,
                MAX(distance) AS max_year
                FROM flights")
```
```{sql connection= con_air}
SELECT MIN(distance) AS min_year,
       MAX(distance) AS max_year
FROM flights
```

```{sql}
--| connection: con_air

SELECT * FROM flights LIMIT 10;
```

Using the Select function to get specific columns from data set
```{sql connection=con_air}
SELECT dep_time, arr_time, flight FROM flights LIMIT 10;
```

Getting average speed in miles per hour for each flight into a variable
```{sql connection= con_air}
SELECT flight, distance/(air_time/60) AS speed FROM flights LIMIT 10;
```

Aggregating on columns. Getting the minimum and maximum air time (in minutes) for flights.
```{sql connection=con_air}
SELECT MIN(air_time) AS min_ar, MAX(air_time) AS max_ar FROM flights;
```

Filtering an observation where conditions are met. Wanting only flights originating from  the JFK airport
```{sql connection=con_air}
SELECT * FROM flights WHERE origin = "JFK" LIMIT 10;
```

FIltering flights not originating in JFK
```{sql connection=con_air}
SELECT * FROM flights WHERE origin != "JFK" LIMIT 10;
```

Finding values in/not in a range.
```{sql connection=con_air}
SELECT * FROM flights WHERE tailnum IN ("N593JB", "N532UA") LIMIT 20;
```
```{sql connection=con_air}
SELECT * FROM flights WHERE month NOT IN (1, 12) AND arr_delay > 120 LIMIT 10;
```

Missing values. Looking at how many observations have no wind gusts
```{sql connection=con_air}
SELECT COUNT(*) AS num_obs FROM weather WHERE wind_gust IS NULL;
```

Group by variables.
```{sql connection=con_air}
SELECT dest, month, day,
       MIN(arr_delay) AS mnd,
       MAX(arr_delay) AS mxd,
       AVG(arr_delay) AS avd
FROM flights
GROUP BY dest, month, day
LIMIT 10;

```

Filtered aggregated values with `having` function. `*` is to select all.
```{sql connection=con_air}
SELECT engines, COUNT(*) AS tot_num
 FROM planes
 GROUP BY engines
 HAVING tot_num < 200;
```

Order By columns when displaying output
```{sql connection=con_air}
SELECT dest, month, day,
       MIN(arr_delay) AS mnd,
       MAX(arr_delay) AS mxd,
       AVG(arr_delay) AS avd
FROM flights
GROUP BY dest, month, day
ORDER BY dest, month DESC, day
LIMIT 10;
```

Using the `distinct` function
```{sql connection=con_air}
SELECT DISTINCT carrier, flight FROM flights ORDER BY carrier, flight LIMIT 10;

```

using the `case when` function
```{sql connection=con_air}
SELECT year, month, day, arr_delay,
       CASE WHEN arr_delay < 0 THEN "Early"
       WHEN arr_delay = 0 THEN "On Time"
       WHEN arr_delay > 0 THEN "Late"
       ELSE "Unknown"
  END AS delay_type
FROM flights LIMIT 10;
```

Subqueries in SQL. Getting all destination codes that start with an M.
```{sql connection=con_air}
SELECT DISTINCT dest FROM flights WHERE dest LIKE "M%"
```
```{sql connection=con_air}
SELECT dest, COUNT(*) as tot_flights
 FROM flights
 WHERE dest IN (SELECT DISTINCT dest FROM flights WHERE dest LIKE "M%")
 GROUP BY dest

```

Relating information between tables in SQL. We're wanting to get the full carrier name not the carrier code. Able to do this with the `JOIN` function.
```{sql connection=con_air}
SELECT carrier, COUNT(*) as tot_flights
 FROM flights
 WHERE month IN (6, 7)
 GROUP BY carrier
```
```{sql connection=con_air}
SELECT al.name AS airline, COUNT(*) as tot_flights
FROM flights AS fl
LEFT JOIN airlines AS al
  ON al.carrier = fl.carrier
WHERE month IN (6, 7)
GROUP BY al.name
```

Saving SQL Query outputs as R objects
```{sql connection=con_air}
--| output.var: tot_airline_flights
SELECT al.name AS airline, COUNT(*) as tot_flights
FROM flights AS fl
LEFT JOIN airlines AS al
  ON al.carrier = fl.carrier
WHERE month IN (6, 7)
GROUP BY al.name
```
```{r}
head(tot_airline_flights, n = 3)
```

Terminated the SQL connection when finished
```{r}
DBI::dbDisconnect(con_air, shutdown = TRUE)
```

